A Benchmark for Consciousness in Large Language Models

### 2024-08-23 tex/paper.tex
update to plans (no longer doing Diplomacy)

Social animals that are widely regarded as conscious share the property that they have developed a theory of mind, and exhibit social behavior. Such animals over the aeons have emergently developed consciousness without other conscious beings training them on every behavior. Although difficult to definitively establish, most theories of consciousness acknowledge that consciousness likely first emerged on earth sometime in the last billion years after the emergence of multicellular live. Furthermore, it has been established that social are more likely to have the capability of self-awareness and demonstrate more signs of consciousness than solitary animals \citep{lei2023sociality}.  We therefore have empirical evidence that selective pressures under the condition of socialization are therefore key to allow for consciousness to emerge.

We are left with the observation that in order for consciousness to emerge, language itself must also emerge in the process without continued examples of natural language, as occurred in the animal kingdom through evolutionary selective pressures.

@@

Next, we introduce the model to the board game diplomacy as was implemented in \citep{fair2022diplomacy} 
the following training scheme to meet the benchmark described above. By playing against itself in a strategic game, it will utilize the reasoning and ability to use words as meaningful tokens. Full-press Diplomacy is a game where effective cooperation, planning, and theory of mind are critical to effective gameplay. As opposed to games like chess or Go, which have been demonstrated to achieve highly sophisticated strategies and seeing many moves ahead, diplomacy does not require seeing as many moves ahead or a deep knowledge of common board states, and instead mostly relies on the ability of players to understand the likely intentions and plans of their opponents, understanding one's own plans, and effectively cooperating to achieve one's goals through written communication. The secondary benefit to using the game is it mimics evolutionary environments of conscious animals, where cooperation, strategy, and theory of mind are critical to achieving reward. Finally, full-press diplomacy has been shown in \citep{fair2022diplomacy} to increase game performance over time with only self-play.

@@

This is why I wanted the model to self-play full-press diplomacy: I wanted it to be rewarded for action that represents internal states accurately, and not for any human-like mimicry. I know self-play can be a very powerful tool to train deep knowledge. All the ACT questions are going to be about what the model plans, prefers, likes, dislikes, knows, doesn't know, how it views others, the structure and breadth of inner experience.

It seems that in terms of understanding natural language (English) without enforcing a model to mimic or infer logically invalid statements, or to hallucinate, there are a limited set of options where training on these tasks or similar tasks has succeeded in the past:
\begin{itemize}
    \item[A:] Training the model on natural language based (e.g. questions written by LLMs or humans) question answering.
    \item[B:] Training on performance on language-involving self play, where self-talk or talk with fellow agents must be constrained to properly formatted English language. E.g. poker, full-press diplomacy, rock paper scissors.
    \item[C:] Training on performance of coding tasks specified in natural language.
    \item[D:] Training on entailment in knowledge databases (provides very large datasets, hundreds of thousands of prompts or even unbounded, with easily constrained input knowledge).
    \item[E:] Training on synthetically generated natural language first order logic tasks (provides very large datasets, hundreds of thousands of prompts or even unbounded, with easily constrained input knowledge).
    \item[F:] Training on many synthetically generated natural language prompts about that model's internal state.
    \item[G:] Training on the degree of certainty of answers to any of the above, or even just simple synthetically generated mathematics questions.
\end{itemize}

@@

In particular, though many possible training schemes may be able to achieve similar results, we recommend a multi-tiered approach to training which begins with item C above, training coding ability with natural language.
We will train a small version of \href{https://github.com/google-deepmind/recurrentgemma}{RecurrentGemma} at only 100 million parameters, and compare its consciousness scores with a similarly sized \href{https://github.com/google/gemma_pytorch}{Gemma} model. The results of both models will be compared side-to-side. If any indications of consciousness appear, it is hypothesized that Gemma would be more likely to exhibit them, although at such small model sizes, and given the exhaustive measures to prevent mimicry and contamination of human thought patterns, it would be extremely surprising if the model exhibited measurable conscious experience.
We intend to first pretrain the model with the \href{https://huggingface.co/datasets/nampdn-ai/tiny-codes}{tinycodes} dataset to increase its reasoning ability and teach it basic natural language skills. The dataset will be filtered for any consciousness-related terms before being used for training and tokenization. Next, the agent will play full-press diplomacy against itself, where communication will be restricted to valid english usage and grammar. The reward will both come from correct communication, and improved performance on diplomacy. To accommodate the small model size, the games will be shortened and fewer players will be used. Reasoning will be required on a "scratchpad" before any communication, to help detect dishonesty.

@@

The first part of the fine tuning is having it understand natural language paragraphs about the external state of the world. In this stage of training we ask it if its knowledge, and any given additional information, validates or invalidate that external state of the world. We can ensure the natural language is at the level of small children, and restrict possible words to further prevent anthropomorphization.

@@

>>>>

Social animals that are widely regarded as conscious share the property that they have developed a theory of mind, and exhibit social behavior. Such animals over the aeons have emergently developed consciousness without other conscious beings training them on every behavior. Although difficult to definitively establish, most theories of consciousness acknowledge that consciousness likely first emerged on earth sometime in the last billion years after the emergence of multicellular live. Furthermore, it has been established that social animals are more likely to have the capability of self-awareness and demonstrate more signs of consciousness than solitary animals \citep{lei2023sociality}.  We therefore have empirical evidence that selective pressures under the condition of socialization are key to allow for consciousness to emerge.

Unfortunately, most conscious animals appear to have limited ability to communicate their internal states to humans. Still, effective language use is an extremely useful tool to communicate internal states and knowledge. A universal trait of social animals is that they communicate their internal states with each other. We hypothesize that in order for consciousness to emerge, communication itself must also emerge in the process. This occurred in the animal kingdom through evolutionary selective pressures, and likely entirely independently for both chordates and mollusks through convergent evolution.

@@

Next, we introduce the model to the board game diplomacy as was implemented in \citep{fair2022diplomacy} 
the following training scheme to meet the benchmark described above. By playing against itself in a strategic game, it will utilize the reasoning and ability to use words as meaningful tokens. Full-press Diplomacy is a game where effective cooperation, planning, and theory of mind are critical to effective gameplay. As opposed to games like chess or Go, which have been demonstrated to achieve highly sophisticated strategies and seeing many moves ahead, high level play of diplomacy does not require seeing as many moves ahead or a deep knowledge of common board states, and instead mostly relies on the ability of players to understand the likely intentions and plans of their opponents, understanding one's own plans, and effectively cooperating to achieve one's goals through written communication. The secondary benefit to using the game is it mimics evolutionary environments of conscious animals, where cooperation, strategy, and theory of mind are critical to achieving reward. Finally, full-press diplomacy has been shown in \citep{fair2022diplomacy} to increase game performance over time with only self-play.

@@

This is why I wanted the model to self-play full-press diplomacy: I wanted it to be rewarded for action that represents internal states accurately, and not for any human-like mimicry. I know self-play can be a very powerful tool to train deep knowledge (just look at success in the boardgame go or DOTA for examples). All the ACT questions are going to be about what the model plans, prefers, likes, dislikes, knows, doesn't know, how it views others, the structure and breadth of inner experience.

It seems that in terms of understanding natural language (English) without enforcing a model to mimic or infer logically invalid statements, or to hallucinate, there are a limited set of options where training on these tasks or similar tasks has succeeded in the past:
\begin{itemize}
    \item[A:] Training the model on natural language based (e.g. questions written by LLMs or humans) question answering.
    \item[B:] Training on performance on language-involving self play, where self-talk or talk with fellow agents must be constrained to properly formatted English language. E.g. poker, full-press diplomacy, rock paper scissors.
    \item[C:] Training on performance of coding tasks specified in natural language.
    \item[D:] Training on entailment in knowledge databases (provides very large datasets, hundreds of thousands of prompts or even unbounded, with easily constrained input knowledge).
    \item[E:] Training on synthetically generated natural language first order logic tasks (provides very large datasets, hundreds of thousands of prompts or even unbounded, with easily constrained input knowledge).
    \item[F:] Training on many synthetically generated natural language prompts about that model's internal state.
    \item[G:] Training on the degree of certainty of answers to any of the above, or even just simple synthetically generated mathematics questions \citep{lin2022teachingmodelsexpressuncertainty}.
.
\end{itemize}

@@

In particular, though many possible training schemes may be able to achieve similar results, we recommend a multi-tiered approach to training which begins with item C above, training coding ability with natural language.
We will train a small version of \href{https://github.com/google-deepmind/recurrentgemma}{RecurrentGemma} at only 100 million parameters, and compare its consciousness scores with a similarly sized \href{https://github.com/google/gemma_pytorch}{Gemma} model. The results of both models will be compared side-to-side. If any indications of consciousness appear, it is hypothesized that Gemma would be more likely to exhibit them, although at such small model sizes, and given the exhaustive measures to prevent mimicry and contamination of human thought patterns, it would be extremely surprising if the model exhibited measurable conscious experience.
We intend to first pretrain the model with the \href{https://huggingface.co/datasets/nampdn-ai/tiny-codes}{tinycodes} dataset to increase its reasoning ability and teach it basic natural language skills. The dataset will be filtered for any consciousness-related terms before being used for training and tokenization.

Next, the agent will play a series of games tuned to improve its ability to predict other agent beliefs and effectively verbally communicate. Communication will be restricted to valid english usage and grammar, and reasonably probable response (a threshold will be applied for communication probability). The reward will both come from correct communication, and improved performance on the social games.

The games will be a variant on diplomacy (TerritoryGame), Rock, Paper Scissors, and a simplified game with the dynamics of Poker. Games are available on github here: \href{https://github.com/morganrivers/consciousness_games}{TerritoryGame}

@@

@@

\bibitem[Lin et al.(2022)]{lin2022teachingmodelsexpressuncertainty}
Lin, S., Hilton, J., \& Evans, O. (2022). Teaching Models to Express Their Uncertainty in Words. \textit{arXiv preprint arXiv:2205.14334}. Retrieved from \url{https://arxiv.org/abs/2205.14334}
